Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-28 paper.tex
Another stab at the discussion

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
%% CQ: Reading this sentence, I thought we were still talking abour simulators, that's what I proposed these changes
Two technological developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing ---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2015scrm}
were necessary to simulate whole chromosomes.

@@

Unit tests are vital for ensuring software quality and correctness, but
they are usually of little value in assessing the statistical properties
of simulations. To validate the correctness of simulation output
we maintain a suite of statistical tests (as of 1.0.0,
217 validation tests, in 6K lines of code). These consist of running many
replicate simulations, and checking the properties of the output
against either analytical results of the output of other simulators
where such results are not available.
For example, simulations of complex demography are validated against \ms,
selective sweeps against \discoal\, and Wright-Fisher simulations
against forwards-time simulations in \SLiM.
This suite of tests is run before every release, to ensure
that statistical errors have not been introduced.

\section*{Discussion}
% Simulation is useful, and it's really easy for a single locus.
% This has lead to a culture of developing lots of simulators
% rather than focusing on one
Simulation has played a key role in the development of coalescent theory
and its use in understanding observed patterns of genetic variation.
The Kingman coalescent can be simulated for $n$ samples in $O(n)$ time
and is straightforward to implement---for example, Hudson provided a fully
functional and efficient simulator as a C source code listing
in a paper appendix~\citep{hudson1990gene}. Because of
this ease, the dominant approach to producing simulators for extensions
to this basic model has been to develop them independently. This is
often seen as a learning opportunity, since those developing extensions
to the coalescent must have an excellent understanding of the basic
model in order to develop extensions to the coalescent.
The coalescent with recombination, however, is \emph{not} straightforward
to implement, and numerous different approaches have been
proposed over the
years~\citep{hudson1983properties,griffiths1997ancestral,wiuf1999recombination,
mcvean2005approximating}. It has only recently become clear that it is
possible to implement the model efficiently, and only through the use
of sophisticated algorithms and data structures~\citep{kelleher2016efficient}.

% Making lots of simulators has given us a fragmented and poor
% quality ecosystem.
The approach of developing ``in-house'' simulators
has lead to a huge proliferation of different programs, with
slightly different feature sets and often incompatible interfaces.
It is a fragmented ecosystem with significant issues with software
quality (see, e.g.,~\cite{yang2014critical}),
and few simulators are actively maintained after publication.
We have seen that \msprime\ (for larger samples at least)
has at worst comparable, but usually far superior
performance to more specialised methods.

% JK: I want to make the point that the established idea that
% developing your own simulation software is a good learning experience
% is pretty questionable these days, and tends to produce a lot of
% crappy unmaintained software. Not sure if the following paragraph
% does the job that well, though.

% There seems to be little point in continuing the huge duplication of effort we
% have seen. It is not possible to achieve the performance levels of \msprime\
% without investing substantially in software engineering. As demonstrated here,
% \msprime\ is extensible and we would argue that developing this simulator as an
% open-source community asset is of more benefit to all. Implementing a
% coalescent simulator may once have been a good way for (e.g.) a student to
% learn about programming and simulation, but it is no longer reasonable to
% expect this. There is simply too much engineering required. For example, as of
% the 1.0.0 release \msprime\ consisted of around 13K lines of C and 11K
% lines of Python, with test suites of 122 C test cases (7K lines of code) and
% 1350 Python test cases (22K lines of code), along with a suite of 217
% statistical validation tests (6K lines of code). It requires a sustained effort
% over years to produce well engineered, high-quality software. It may reasonably
% be asked if students would learn more from contributing to an existing project
% with established high-quality engineering standards, or from producing their own
% software in a self-taught manner.

% A significant advantage to simulating the coalescent with recombination
% backwards in time using Hudson's algorithm rather than
% left-to-right
% approaches~\citep{wiuf1999recombination,chen2009fast,staab2015scrm}
% is the ease with which additional processes can be integrated.
% We have seen that processes such as gene conversion, complex
% demography, instantaneous bottlenecks, multiple merger caolescents
% and selective sweeps have been integrated seamlessly into
% the simulations. Incorporating any of these extensions into
% either the exact or approximate left-to-right processes would require
% complex derivations. Because we are going backwards
% in time, processes can be modelled as competing exponentials
% using Gillespie's algorithm \jkcomment{I think?},
% and therefore in principle we can
% add as many extra processes as we like.
% \jkcomment{so, backwards in time
% sims in msprime is extensible for the future.}

>>>>

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.
Two technological developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing ---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2015scrm}
were necessary to simulate whole chromosomes.

@@

Unit tests are vital for ensuring software quality and correctness, but
they are usually of little value in assessing the statistical properties
of simulations. To validate the correctness of simulation output
we maintain a suite of statistical tests (as of 1.0.0,
217 validation tests, in 6K lines of code). These consist of running many
replicate simulations, and checking the properties of the output
against either analytical results of the output of other simulators
where such results are not available.
For example, simulations of complex demography are validated against \ms,
selective sweeps against \discoal, and Wright-Fisher simulations
against forwards-time simulations in \SLiM.
This suite of tests is run before every release, to ensure
that statistical errors have not been introduced.

\section*{Discussion}

% Simulation software is historically a total shitshow

Simulation is a vital tool in population genetics, but the approach to
developing simulation software has historically been dysfunctional. A huge
proliferation of tools have been published (the references here are not
exhaustive) and only a small minority of these are actively developed and
maintained today. The ecosystem is highly fragmented, with numerous different
ways of specifying parameters and representing results, and there are
significant software quality issues at all stages. This is unsurprising, since
the majority of simulation software development is performed by students, who
are themselves learning to code, often without formal training in software
development. Although developing a simulator is certainly a learning
opportunity, it is also an opportunity to learn poor development practices. A
large number of unmaintained tools with mostly overlapping features represents
a profound waste of effort, and it is worth considering what could have been
achieved if even a small fraction of this effort had been invested in shared
community infrastructure.

% THERE IS A BETTER WAY!!!
Scientific software is vital apparatus, and must be engineered
to a high quality if we are to trust its results. There is a growing
realisation across the sciences [citations] that investing in
shared community infrastructure produces better results than a
proliferation of individually maintained tools, allowing scientists
to focus on their specific questions rather than software engineering.
\texttt{Msprime} 1.0 is the result of such a community process,
with features added by motivated users, taking advantage of the
established development practices and infrastructure.
% I'm going out on a limb here - do folks agree?
Most contributors would agree that this is also a learning experience,
allowing them to improve their software development skills in a
welcoming community environment and being exposed to
state-of-the-art infrastructure. We hope that users who find that
features they require are missing will continue to contribute to
\msprime, leading to a community project that is both high-quality
and sustainable in the long term.

\jkcomment{TODO either need another paragraph linking these up, or we just
 drop this last one?}
